---
title: "Day 1 session 3 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
editor_options: 
  chunk_output_type: console
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}


# install.packages("gapminder")

download.file("https://www.openintro.org/data/rda/resume.rda", "resume.rda") # hypo test proportions

rworkshop::download_data("gingko_RCT.rda") # hypo test means

rworkshop::download_data("amazon.rda")   # hypo test correlation



rworkshop::download_data("IPED_salaries_2016.rda")  # linear modeling

rworkshop::download_data('downloading.csv') # one-way ANOVA

rworkshop::download_data("freshman-15.txt") # random effects model

```







```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

library(dplyr)


#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

# hide all plot output - useful for printing the code
# knitr::opts_chunk$set(fig.show='hide')

set.seed(123)

```






$\\$



## Overview


 * Hypothesis test for:
   - means
   - proportions
   - correlation
   
  * Permutation test for correlation

  * Simple linear regression



$\\$



## Part 1: parametric hypothesis test



$\\$



### Part 1.1:  Parametric hypothesis test for means


Does Gingko change cognitive abilities? 


$H_0: \mu_{gingko} - \mu_{control} = 0$
$H_A: \mu_{gingko} - \mu_{control} \ne 0$



```{r gingko}

load("gingko_RCT.rda")


# plot the data
boxplot(gingko, placebo, 
        names = c("Gingko", "Placebo"),
        ylab = "Memory score")


# run a t-test
t.test(gingko, placebo)
     

```





$\\$





### Part 1.2:  Parametric hypothesis test for proportions


Do resumes with female sounding names get fewer call backs than resumes with male sounding names? 


$H_0: \pi_{female} = \pi_{male}$ 
$H_A: \pi_{female} \ne \pi_{male}$ 



The function `prop.test()` takes a table with two columns giving the counts of successes and failures, respectively.


```{r resumes}


load("resume.rda")


# get a table showing call backs for male and female sounding names
count_table <- table(resume$gender, resume$received_callback)
     

# visualize the data
barplot(t(prop.table(count_table, margin = 1)))


# run a hypothesis test
prop.test(count_table)


```





$\\$


### Part 1.3:  Parametric hypothesis test for correlation

Do books with more pages cost more? 


$H_0: \rho = 0$

$H_0: \rho > 0$




```{r books}

load("amazon.rda")


# extract the number of pages and price
num_pages <- amazon$NumPages
price <- amazon$List.Price


# plot the data
plot(num_pages, price)


# run the hypothesis test
cor.test(num_pages, price)


```





$\\$





## Part 2: Permutation test for correlation


```{r permtest_corr}


# calculate the observed statistic
obs_stat <- cor(num_pages, price)


# create the null distribution 
null_dist <- NULL
for (i in 1:10000) {
  
  curr_null_stat <- cor(num_pages, sample(price))
  null_dist[i] <- curr_null_stat
  
}



# visualize the null distribution 
hist(null_dist, breaks = 100)
abline(v = obs_stat, col = "red")


# Calculate the two-tailed p-value
(pval <- mean(null_dist >= obs_stat))

  

```





$\\$





## Part 3: simple linear regression 


Let's look at the relationship between how much faculty get paid and the size of a school's endowment. 



$\\$



### Part 3.1: Visualize the data



```{r}

library(dplyr)

load("IPED_salaries_2016.rda")


# getting only data for assistant professors - ignore this for now and we will discuss this tomorrow
assistant_data <- IPED_salaries |>
  filter(endowment > 0,  rank_name == "Assistant") |>
  mutate(log_endowment = log10(endowment)) |>
  mutate(log_enroll = log10(enroll_total)) |>
  select(school, salary_tot, endowment, log_endowment, enroll_total, log_enroll, salary_men, salary_women) |>
  na.omit() |>    # get only the cases that do not have missing data
  arrange(endowment)



# visualize the data
plot(assistant_data$endowment, assistant_data$salary_tot)


# visualize log10 transformation of the endowment
plot(assistant_data$log_endowment, assistant_data$salary_tot)



```




$\\$






### Part 3.2: Fitting a linear regression model


Let's now fit a linear model to our data to predict faculty salaries from the `log10` of the endowment. 



```{r linear_fit}


# fit a regression model
lm_fit <- lm(salary_tot ~ log_endowment, data = assistant_data)


# plot the data and add the regression line to the plot
plot(assistant_data$log_endowment, assistant_data$salary_tot)
abline(lm_fit, col = "red")


# look at the coefficients (beta hat's)
coef(lm_fit)


```





$\\$




### Part 3.3: Hypothesis tests for regression coefficients using parametric methods


Let's use the t-statistic and t-distribution to do parametric inference on
whether the regression offset $\beta_0$, and the regression slope $\beta_1$, are
different than 0. This means we are testing the hypotheses:

$H_0: \beta_0 = 0$
$H_A: \beta_0 \ne 0$


and


$H_0: \beta_1 = 0$
$H_A: \beta_1 \ne 0$



```{r reg_hypothesis_tests}


# Assessing if the regression coefficients are statistically significant
summary(lm_fit)


```



$\\$




### Part 3.3: Calculating confidence intervals for regression coefficients



```{r reg_CIs}

# get a confidence interval for the regression coefficients
confint(lm_fit, level = .95)


```




$\\$





### Part 3.4: Regression diagnostics


```{r regression_diagnostics}


# view the regression diagnostic plots
par(mfrow = c(2, 2))

plot(lm_fit)



```






$\\$






## Part 4: Multiple regression 


Let's look at multiple regression where we have more than 1 predictor


$\\$



### Part 4.1: Multiple regression


Let's build a regression model where we try to predict salary as a function of log endowment and log enrollment.


```{r}


par(mfrow = c(1, 1))


# plot the relationship between log enrollment and salary
plot(assistant_data$log_enroll, assistant_data$salary_tot)


# build the regression model
(lm_fit_endow_enroll <- lm(salary_tot ~ log_endowment + log_enroll, data = assistant_data))

summary(lm_fit_endow_enroll)



```




$\\$





### Part 4.2: Test for comparing nested models


When we have nested models, we can use an ANOVA test based on the F-statistic to
assess if adding additional explanatory variables leads to a model that can
account for more of the variability in a response variable.


A Model 1 is nested in a Model 2 if the parameters in Model 1 are a subset of
the parameters in Model 2. Here, our model using only the endowment as the
explanatory variable is nested within in the model that uses endowment and
enrollment as explanatory variables. Let's uses the `anova()` function to test
if adding the enrollment explanatory leads to a statistically significant
increase in the amount of variability that can be accounted for.


```{r}

anova(lm_fit, lm_fit_endow_enroll)

```




$\\$



