---
title: "Day 2 session 4 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
# rworkshop::update_installed_packages()


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * String manipulation



$\\$






## Part 1: String manipulation using stringr


$\\$


##### Part 1.1: Base R and stringr


```{r}

library(stringr)







```



$\\$


##### Part 1.2: Trimming and padding a string


```{r}








```



$\\$



##### Part 1.3: Extracting a piece of a string


```{r}


fruits <- c("apple", "pineapple", "Pear", "orange",  "peach", "banana")




```



$\\$



##### Part 1.4: Concatenating strings


```{r}







```



$\\$





##### Part 1.5: Download an article



```{r}


#base_name <- "https://www.washingtonpost.com/politics/2021/11/16/white-house-says-democratic-infighting-has-hurt-biden/"
#article_name <- "infighting.html"
#full_name <- str_c(base_name) #, article_name)


base_name <- "https://www.nytimes.com/2021/11/17/us/"
article_name <- "catholic-bishops-biden-communion.html"
full_name <- str_c(base_name, article_name)


# download.file(full_name, article_name)


# viewer <- getOption("viewer")
# viewer(article_name)




```



$\\$


##### Part 1.6: string length



```{r}











```



$\\$



##### Part 1.7: string replace all


```{r}








```




$\\$


##### Part 1.8: String split


```{r}








```




$\\$


##### Part 1.9: Extracting strings


```{r}








```



$\\$


##### Part 1.10: Detecting a string


```{r}






```



$\\$





## Part 2: Regular expressions


We can use regular expressions to do a lot more complex string matching!




$\\$





### Part 2.1: Start and end of strings

Using regular expression we can match the start and end of words using: 

* Match the start of a string with ^
* Match the end of a string with $
* [Aa] detects a lower or upper case A



```{r start_end}

# detect all fruits that end with e



# detect all fruits that start with lower or upper case P



```





$\\$





### Part 2.2: Matching characters 


Using regular expression match characters using: 


* Match any character with .
* Match 0 or more characters with *
* Match 1 or more characters with +



```{r any_characater}

# detect "mess", "mass", and "miss" from the following string




# detect all the strings from the xz_string




# detect all but the first string from the xz_string




```




$\\$





### Part 2.3: Matching anything in a range

* [ ] means match anything in the range inside the braces 
* If the ^ appears inside square braces it means not


```{r range_detect}

# detect all fruits that start with a vowel


# detect all these strings




```





$\\$




### Part 2.4: Detecting repeats


* {num} means repeat the preceding sequence num times
* () groups things together, useful in combination with {}

```{r repeats}

# detect all fruits with an in them 



# detect all fruits with anan in them 



```





$\\$





### Part 2.5: Example - extracting phone numbers

Can you extract all the phone numbers from these strings? 


```{r phone_numbers}

input_strings <- c(
	"apple", 
  	"219 733 8965", 
  	"329-293-8753", 
  	"Work: 579-499-7527",
	"Home: 543.355.3679"
)








```



$\\$





### Part 2.6: Escape sequences


In regular expressions a period (.) means any character. So how can you detect if a period is in a string? 

Escape sequences in R start with two slashes \\ and cause the next character to be treated literally rather than as a special character
To match a period we use   `\\.`            [.] also works
To match a `$` symbol we use  `\\$`



```{r escape_sequences}

# Extract the amounts of money and dollar sign from this string  (use str_extract_all) 
the_string  <-  c("Joanna has $100 and Chris has $0") 




```





$\\$





### Part 2.7: Special characters


Other special characters are also designated by using a double slash first
\\s   space
\\n   new line     or also   \\r
\\t   tab


```{r special_characters}

# get 6 characters prior to the end of a line in the_article



# extract all lines that have html tags at the end of the line
# i.e., that have </ > at the end of a line




```


