---
title: "Day 2 session 4 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
# rworkshop::update_installed_packages()

# rworkshop::download_data("moby_dick.txt")

```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * String manipulation



$\\$






## Part 1: String manipulation using stringr




$\\$




### Part 1.1: Base R and stringr


```{r}

tolower("Hey")


library(stringr)


str_to_lower("STOP YELLING")




```





$\\$





### Part 1.2: Trimming and padding a string


```{r}


str_trim("    What a mess      ")
str_to_lower("STOP YELLING")


str_pad("Let's make it messier", 50, "right")
str_pad(1:11, 3, pad = 0)   # useful for adding leading 0’s




```





$\\$





### Part 1.3: Extracting a piece of a string


```{r}

str_sub("What a mess", 6, 11)

fruits <- c("apple", "pineapple", "Pear", "orange",  "peach", "banana")
str_sub(fruits, 2, 4)


```





$\\$





### Part 1.4: Concatenating strings


```{r}

str_c("What", "a", "mess", sep = " ")

vec_words <- c("What", "a", "mess")
str_c(vec_words, collapse = " ")


```






$\\$






### Part 1.5: Download an article



```{r}


base_name <- "https://www.foxnews.com/politics/biden-deliver-unscheduled-speech-capitol"
article_name <- "politics.html"
download.file(base_name, article_name)




# viewer <- getOption("viewer")
# viewer(article_name)




```






$\\$





### Part 1.6: string length



```{r}


str_length("What a mess")
 

article_size <- file.info(article_name)$size    # size of the article in bytes

# read the whole article as a single string
the_article <- readChar(article_name, article_size)


str_length(the_article)     # size of the article as a string





```




$\\$





### Part 1.7: string replace all


```{r}

str_replace("String", "old", "new")


article2  <- str_replace_all(the_article,  "Biden", "Sleepy Joe")

write(article2, "sleepy_article.html")

#viewer("sleepy_article.html")



```




$\\$





### Part 1.8: String split


```{r}


str_split("String",  "split pattern")


list_of_strings <- str_split("What a mess",  " ")
vector_of_strings <- unlist(list_of_strings)
vector_of_strings[3]

article_vec <- unlist(str_split(the_article, " "))


```




$\\$





### Part 1.9: Extracting strings


```{r}



str_extract("String", "pattern")

str_extract(fruits,  "apple")



```





$\\$





### Part 1.10: Detecting a string


```{r}

str_detect("String", "pattern")

str_detect(fruits,  "apple")

# can you tell how many times Biden was mentioned in the article? 
sum(str_detect(article_vec, "Biden"))


```



$\\$




## Part 2: Regular expressions


We can use regular expressions to do a lot more complex string matching!




$\\$





### Part 2.1: Start and end of strings

Using regular expression we can match the start and end of words using: 

* Match the start of a string with ^
* Match the end of a string with $
* [Aa] detects a lower or upper case A



```{r start_end}

# detect all fruits that end with e
str_detect(fruits, "e$")


# detect all fruits that start with lower or upper case P
str_detect(fruits, "^[Pp]")


```





$\\$





### Part 2.2: Matching characters 


Using regular expression match characters using: 


* Match any character with .
* Match 0 or more characters with *
* Match 1 or more characters with +



```{r any_characater}

# detect "mess", "mass", and "miss" from the following string
mss_strings <- c("mess", "best", "mass", "kiss", "miss")
str_detect(mss_strings, "m.ss")


# detect all the strings from the xz_string
xz_strings <- c("xz", "xyz", "xyyz", "xyyyz")   
str_detect(xz_strings , "xy*z")


# detect all but the first string from the xz_string
str_detect(xz_strings, "xy+z")



```




$\\$





### Part 2.3: Matching anything in a range

* [ ] means match anything in the range inside the braces 
* If the ^ appears inside square braces it means not


```{r range_detect}

# detect all fruits that start with a vowel
str_detect(fruits, "^[a-o]")

# detect all these strings
ch_strings <- c("chimp", "champion", "chomp")
str_detect(ch_strings, "ch[aio]mp")


```





$\\$




### Part 2.4: Detecting repeats


* {num} means repeat the preceding sequence num times
* () groups things together, useful in combination with {}

```{r repeats}

# detect all fruits with an in them 
str_detect(fruits, "(an){1,}")

# detect all fruits with anan in them 
str_detect(fruits, "(an){2}")


```





$\\$





### Part 2.5: Example - extracting phone numbers

Can you extract all the phone numbers from these strings? 


```{r phone_numbers}

input_strings <- c(
	"apple", 
  	"219 733 8965", 
  	"329-293-8753", 
  	"Work: 579-499-7527",
	"Home: 543.355.3679"
)

phone_expression <- "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"

str_extract(input_strings, phone_expression)



```



$\\$





### Part 2.6: Escape sequences


In regular expressions a period (.) means any character. So how can you detect if a period is in a string? 

Escape sequences in R start with two slashes \\ and cause the next character to be treated literally rather than as a special character
To match a period we use   `\\.`            [.] also works
To match a `$` symbol we use  `\\$`



```{r escape_sequences}

# Extract the amounts of money and dollar sign from this string  (use str_extract_all) 
the_string  <-  c("Joanna has $100 and Chris has $0") 
str_extract_all(the_string, "\\$[0-9]{1,}")

```





$\\$





### Part 2.7: Special characters


Other special characters are also designated by using a double slash first
\\s   space
\\n   new line     or also   \\r
\\t   tab


```{r special_characters}

# get 6 characters prior to the end of a line in the_article
str_extract_all(the_article, ".{6}\\n") 


# extract all lines that have html tags at the end of the line
# i.e., that have </ > at the end of a line
end_tag <- str_extract_all(the_article, "</[A-z]{1,}>\\n")
lapply(end_tag, str_replace, "\n", "")


```




