---
title: "Day 2 session 2 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}


# install.packages("gapminder")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

library(dplyr)


#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview


 * Plotting with ggplot



$\\$








## Part 1: Visualizations with ggplot


We can use the ggplot2 package, which is part of the tidyverse, to create much
nicer looking graphics than using base R graphics. The ggplot2 library is
modeled on Leland Wilkinson's "grammar of graphics" which creatse
graphics from a combination of basic visual elements.

In the exercises below, we will learn how to use ggplot using the motor trends
cars data set (mtcars) that comes with base R installation, and also the
gapminder data.


A few resources to learn more about ggplot are: 

- [R for Data Science](https://r4ds.had.co.nz/)
- [The ggplot cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)





$\\$





### Part 1.1: Scatter plots


Let's create plots of the number of miles per gallon (mpg) cars get as a
function of the weight of the car.


```{r scatter_plot}


# install.packages("ggplot2")

library(ggplot2)



# base R





# ggplot 






# ggplot - mapping in the geom





```




$\\$





### Part 1.2: Adding labels to plots


We can add labels to the plots using the `labs()` function. Arguments to `labs()` function include:
- `x`: the label on the x-axis (you can also use the `xlab()` function)
- `y`: the label on the y-axis (you can also use the `ylab()` function)
- `title`: the title of the plot  (you can also use the `ggtitle()` function)
- `subtitle`: the title of the plot  (you can also use the `ggtitle()` function)

If you just want to add x and y labels or a title, you can also use the
`xlab("label1")`, `ylab("label2")` and/or the `ggtitle()` functions.



```{r labels}












```



Remember, if you don't want exes [label your axes](https://xkcd.com/833/)



$\\$




### Part 1.3: Adding annotations to plots

We can add annotations to plots using the `annotate("text", x = , y = , label = )` function.



```{r annotations}










```







$\\$





### Part 1.4: More aesthetic mappings


We can use other aesthetic mappings beyond position including:
- `color`: different color scales are used for quantitative and categorical (factor) data 
- `shape`: should be categorical data
- `size`: should be quantitative data



```{r mappings}


# add color based on the transmission type (is automatic or not)




# it is better to treat the transmission type as a categorical variable?





# can also try mapping transmission type to shape or size 









```



Question: When adding the variable automatic/manual transmission (am) to the
scatter plot mpg vs. weight, do you think it is best to map am on to...?

a. color
b. shape
c. size




$\\$





### Part 1.5: Attributes vs. Aesthetics


Setting an **aesthetic mapping** maps a variable to a **glyph property**. This
is done **inside** the `aes()` function.

Setting an **attribute** set a glyph property to a **fixed value**. This is done
**outside** the `aes()` function.



```{r aesthetics_vs_attributes}


# setting an aesthetic mapping




# setting an attribute





```




$\\$





### Part 1.6: Facets


Beyond comparing variables based on aesthetic mappings, you can compare
categorical variables by splitting a plot into subplots, called facets, using
`facet_wrap()`


```{r facets}


# separate subplots for the two transmission types





# One can also do facets in two dimensions





```




$\\$





### Part 1.7: Overplotting


Sometimes points overlap making it hard to estimate the number of points at a
particular range of values.

We can control the transparency of points by changing their alpha values.



```{r transparency}

library(gapminder)


# a lot of overplotting




# changing the transparency levels makes it a bit easier to see how many points are at a given x, y location




```





$\\$





### Part 1.8: Changing scales


Each visual attribute that has an aesthetic mapping has a default scales. We can
change the scales used for each mapping using functions that start with
`scale_`.

For example, we can change the x-scale from liner to logarithmic using
`scale_x_continuous(trans='log10')`. Likewise we can change the color scale
using `scale_color_manual()`.



```{r scales}

# changing the scale on the x-axis





# mapping continents to colors, and adding my own color scale





```






$\\$






### Part 1.9: geoms


We can use different geoms to create other types of plots.



```{r geoms1}


# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_line()





# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_col()





# Let's MPG as a function of weight using the names of cars rather than just points for each car





# Let's plot a histogram of the weights of cars




# Let's create a boxplot of the weights of cars



# Let's create a side-by-side boxplot of the weights of cars depending on the number of cylinders the engine has




```





$\\$





### Part 1.10: geoms continued

Violin and Joy plots are other ways to view distributions of data



```{r geoms2}

# violin plot 




library("ggridges")

# joy plot




```


Question: Can you figure out where the name "joy plot" comes from? 





$\\$






### Part 1.11: geoms continued


We can also have multiple geom layers on a single graph by using the + symbol
E.g   `ggplot(…) + geom_type1() + geom_type2()`



```{r geoms3}

# Create a scatter plot of miles per gallon as a function of weight and then add a smoothed line using geom_smooth() and a vertical lines using geom_vline()







```





$\\$





### Part 1.12: Themes


We can also use different themes to change the appearance of our plot.


```{r themes}


# Add theme_classic() to our plot



# install.packages("ggthemes")


# library(ggthemes)







```




$\\$




### Part 1.13: Exeriment yourself


Try to create some interesting visualizations from either a data set we have used in the class, or a new data set you found.




```{r experiment}










```






$\\$







## Part 2: Additional ggplot themes and features



$\\$




### Part 2.1: Color scales using scale_color_brewer


Let's use the gapminder data to construct a visualization using this grammar that has:

1. A basic frame constructed from called the `ggplot()` function

2. A geom of points where:
  - `gdpPercap` is mapped to the x position
  - `lifeExp` is mapped to the y position
  - `continent` is mapped to color

3. Transform the scales so that 
  - The x values are on a $log_10$ scale
  - **The color scale using a different qualitative color brewer pallet**


```{r ggplots1}

library(ggplot2)
library(gapminder)


ggplot(data = gapminder) +                                                 # create the basic frame
	geom_point(aes(x = gdpPercap, y = lifeExp, color = continent))  +        # add a geom with 3 aesthetic mappings 
  scale_x_continuous(trans = "log10") + 			                             # change the scales
  scale_color_brewer(type = "qua", palette = 2)                     


```






$\\$






### Part 2.2: ggplot themes


Let's continue to use the gapminder data to construct a visualization using this grammar that has:

1. The gapminder data filtered to only have data from France, Haiti and the United States

2. A geom of line where:
  - `year` is mapped to the x position
  - `lifeExp` is mapped to the y position

3. Faceting based on the country

4. A second glyph layer with a red vertical line at the year 1969

5. **The Wall Street Journal theme from the ggthemes package**



```{r ggplots2}

library(dplyr)
library(ggthemes)


gapminder |>
  filter(country %in% c("France", "Haiti", "United States")) |>
ggplot(aes(year, lifeExp)) + 
  geom_line() + 
	facet_wrap(~country)  +        # add a geom with 3 aesthetic mappings 
  geom_vline(xintercept = 1969, col = "red") + 
  theme_wsj()



```




$\\$





### Part 2.3: geom_text()


We can also create:
- Text as a glyph using `geom_text()`
- Annotations using `annotate(geom_type, geom_properties)` 
- Manually modify the theme using `theme()` with different arguments


Let's create a plot where:

1. It only has data from 2007

2. A text glyph with the following properties:
  - `gdpPercap` is mapped to the x position
  - `lifeExp` is mapped to the y position
  - `continent` is mapped to color
  - `country` is mapped to the text label
  
3. Let's use a log 10 scale

4. Let's turn off the legend by modifying the theme

5. Let's add a text annotation with the following attributes
  - x = 600
  - y = 80
  - label = "this plot is messy :("


```{r ggplot3}

gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, label = country, color = continent)) + 
  geom_text() + 
  scale_x_continuous(trans = "log10") + 
  theme(legend.position="none") + 
  annotate("text", x = 600, y = 80, label = "this plot is messy :(") 

```




$\\$







### Part 2.4: Plotly 


The plotly package can be used to create interactive visualizations. 

Let's recreate an interactive version of the gapminder data from 2007 using a `geom_point()` and:

  - `gdpPercap` is mapped to the x position
  - `lifeExp` is mapped to the y position
  - `continent` is mapped to color
  - `country` is mapped to the name

  
We will save our plot to an object `g` and then we can use `ggplotly(g)` to create an interactive visualization. 



```{r, plotly}

library(gapminder)
library(plotly)


g <- gapminder |> 
  filter(year == 2007) |>  	
  ggplot(aes(x = gdpPercap, y = lifeExp, col = continent, name = country)) + 	  
  geom_point()   

# ggplotly(g)


```





$\\$






### Part 2.5: emojis


Everyone loves emojis! 

Let's use the motor trends car data to plot miles per gallon (mpg) as a function of weight using the proper glyph.



```{r, emoji}


# devtools::install_github("dill/emoGG")
library(emoGG)
ggplot(mtcars, aes(wt, mpg)) + 
   geom_emoji(emoji="1f697")



# another option...
#install.packages("emojifont")
# 
# library(emojifont)
# 
# load.emojifont('OpenSansEmoji.ttf')
# 
# 
# ggplot(mtcars, aes(x = wt, y = mpg, label = emoji("car"), col = factor(cyl))) +  	  	
# 	geom_text(family="OpenSansEmoji", size=6)  +    
# 	xlab("Weight") +   
# 	ylab("Miles per Gallon") + 
# 	theme_classic()



```






$\\$






### Part 2.6: animations


We can crate animated gifs too...



```{r, animations}


#install.packages('png')
#install.packages('gifski')
#install.packages('gganimate')


# library(gganimate)
# 
# 
# ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
#   geom_point(alpha = 0.7, show.legend = FALSE) +
#   scale_colour_manual(values = country_colors) +
#   scale_size(range = c(2, 12)) +
#   scale_x_log10() +
#   facet_wrap(~continent) +
#   # Here comes the gganimate specific bits
#   labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
#   transition_time(year) +
#   ease_aes('linear')



```






$\\$
