---
title: "Day 1 session 2 notes and code"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





$\\$





<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

rworkshop::download_data("daily_bike_totals.rda")
rworkshop::download_data("profiles_revised.csv")


```





## Part 1: Examining categorical data

Categorical variables take on one of a fixed number of possible values 

For categorical variables, we usually want to view:

 * How many items are each category    or
 * The proportion (or percentage) of items in each category




$\\$




### Part 1.1: Count tables


Let's examine categorical data by looking at drinking behavior on OkCupid


```{r categorical_1}

# read the data into a data frame
profiles <- read.csv("profiles_revised.csv")


# Get information about drinking behavior



# Create a table showing how often people drink




```





$\\$



### Part 1.2: Relative frequency tables



We can create a relative frequency table using the function: `prop.table(my_table)` 

Can you create a relative frequency table for the drinking behavior of the people in the okcupid data set in the R chunk below? 


```{r categorical_2}

# create a relative frequency table



```





$\\$



### Part 1.3: bar plots (no pun intended)




We can plot the number of items in each category using a bar plot: barplot(my_table) 

Can you create a bar plot for the drinking behavior of the people in the okcupid data set? 

```{r categorical_3}

# create a bar plot




```





$\\$



### Part 1.4: pie charts



Is there a problem with using the bar plot function without any of the extra arguments? 

[XKCD illusterates the point](https://xkcd.com/833/)

Can you figure out how to fix your plot? 

We can also create pie charts using the pie function

```{r categorical_4}

# create a pie chart


```



Some pie charts [are more informative than others](http://robslink.com/SAS/democd91/pyramid_pie.htm)




$\\$




[There are other websites/apps for dating as well](https://i.kym-cdn.com/photos/images/original/001/293/992/d80.jpg)





$\\$






## Check out the (older) OkCupid data analysis blog


For example, here is an interesting article on [The Big Lies People Tell In Online Dating](https://theblog.okcupid.com/the-big-lies-people-tell-in-online-dating-a9e3990d6ae2). 







$\\$





## Part 2: Examining quantiative data


$\\$


### Part 2.1:  Statistics for central tendency 

There are several summary statistics useful for describing quantitative data
such as the mean and the median. Use the `mean()` and `median()` functions to
extract measures of the central tendency for OkCupid user's heights.


```{r quantitative_1}

# calculate the mean of the heights
mean(profiles$height)

```




What went wrong? 


We can ignore missing values using the `na.rm = TRUE` argument


```{r quantitative_2}

# try again to calculate the mean and median of the heights
mean(profiles$height, na.rm = TRUE)
median(profiles$height, na.rm = TRUE)

```





$\\$






### Part 2.2:  Histograms


We can plot histograms of heights using the `hist()` function. 

```{r quantitative_4}

# plot a histogram



```




$\\$




We can add lines to our plots using the `abline()` function. For example `abline(v = 60)` 
would add a vertical line at the value of 60. Can you add a vertical line
at the average OkCupid user's height?



```{r quantitative_5}

# add a vertical line to a histogram 




```




$\\$





### Part 2.3:  Boxplots



Boxplots visually show a version of a 5 number summary (min, Q1, median, Q3,
max). We can create boxplots using the `boxplot()` function.

Create a boxplot of OkCupid user's heights.


```{r quantitative_6}

# create a box plot



```

If there are extreme outliers in a plot we need to investigate them. If they are
errors we can remove them, otherwise we need to take them into account.



$\\$






### Part 2.4:  Scatter plots and time series



Let's now look at data from CitiBike in New York City. How many cases are there
any how many variables? What does each case correspond to?


```{r quantitative_7}

load("daily_bike_totals.rda")


```





$\\$




Scatter plots show the relationships between two quantitative variables. We can
use the `plot(x, y)` function to create scatter plots. Create a scatter plot of
the maximum temperature as a function of the minimum temperature. Also create a
scatter plot of the number of trips as a function of the date.



```{r quantitative_8}

# plot a time series



```



$\\$





## Part 3: Generating random data and probability functions 


$\\$



### Part 3.1: Generating random data


There are a series of R functions that can generage random data. They are all start with the letter r. 

 * `rnorm()`: normal 
 * `runif()`: uniform 
 * `rexp()`: exponential
 * `rbinom()`: binomial 


Let's plot 1000 randomly generated points from a normal distribution.


```{r random_data}

# plot 1,000 random values from a normal distribution 



```



$\\$




### Part 3.2: Calculating probabilities



We can also get the probability a random variable X will be less than a
particular value x; i.e., P(X < x) using functions that start with the letter p.


 * `pnorm()`: normal 
 * `punif()`: uniform 
 * `pexp()`: exponential
 * `pbinom()`: binomial 


What is the probability we will get a value less than 0 from a standard normal distribution? 


```{r probability}

# calculate P(X < 0) from a standard normal distribution



```


$\\$





### Part 3.2: Getting quantile values



We can get quantiles using functions that start with the letter q. 


 * `qnorm()`: normal 
 * `qunif()`: uniform 
 * `qexp()`: exponential
 * `qbinom()`: binomial 


What is the quantile value q such that area 97.5% of the area under a normal distribution is less than q?


```{r quantiles}

# get a quantile q: P(X < p) = .975  from a standard normal distribution


```


$\\$




## Part 4: For loops  


$\\$



### Part 4.1: For loop basics



For loops are useful when you want to repeat a piece of code many times under similar conditions

Print the numbers from 1 to 50...


```{r}

# create a simple for loop 




```



$\\$




### Part 4.2: For loop working with vectors



For loops are particular useful in combination with vectors that can store the results.

Create a vector with the squares of the numbers from 1 to 50. 



```{r}

# create a loop that creates a vector with the squares of the numbers from 1 to 50. 






```




$\\$




### Part 4.3: Simulating a sampling distribution


We can use a for loop to simulate a sampling distribution.



```{r sampling_dist}


# create a sampling distribution of the mean using data from a uniform distribution




# plot a histogram of the sampling distribution of these means 





```



$\\$




## Part 5: Writing functions


So far we have used several different functions that come with R. We can write
our own function using the `function()` function.




### Part 5.1: Writing a simple function


Let's write a function called `cube_it()` that takes a value `x` and returns `x` cubed.



```{r writing_functions}


# the square root function
sqrt(49)



# write a function that takes a value to the third power








```






$\\$






### Part 5.2: Additional function arguments


We can specify additional arguments which can also have default values. 

Let's write a function called `power_it` which takes a value `x` and a value
`pow` and returns `x` to the pow power. Let's also have the default value of the
`pow` argument be 3.


```{r writing_functions2}


# a function that takes a value to any power








```





$\\$








